Implementation and verification of a data structure using dependent types

### 2022-05-02 paper/main.tex
Plot time vs radix

@@

@@

@@

We know that for each natural number $n$ there are $d \in \mathbb{N}$ and $n_{d - 1}, n_{d - 2}, ..., n_0, n_1 \in \{0, 1\}$ such that $n = n_{d - 1} 2^{d - 1} + n_{d - 2} 2^{d - 2} + ... + n_1 2^1 + n_0 2^0$. The function \coqinline{to_digits : forall (d n : nat), list nat} gives the list of length \coqinline{d} consisting of these binary digits of \coqinline{n} in big-endian ordering: for example, \coqinline{to_digits 8 73 ~> [0; 1; 0; 0; 1; 0; 0; 1]} (where ``\nospaceevaluatesto'' means ``evaluates to''). The inverse operation is \coqinline{of_digits : forall (d : nat) (nl : list nat), nat}: for example, \coqinline{of_digits 8 [0; 1; 0; 0; 1; 0; 0; 1] ~> 73}.

@@

Here is an example of a digit list for $n = 5$, so $d = 3$, $n_2$ = 1, $n_1 = 0$, $n_2 = 1$, which stores values 0, 1, 2, 3, 4:

@@

So in the case $i_{d - 1} = n_{d - 1} = 0$ (that is, \coqinline{il} starts with 0 and there is no tree in this position) and in the case $i_{d - 1} = n_{d - 1} = 1$ (that is, \coqinline{il} starts with 1 and there is a tree in this position), we recurse to our tail. Otherwise, in the case $i_{d - 1} = 0$ and $n_{d - 1} = 1$ (that is, \coqinline{il} starts with 0 and there is a tree in this position), we ask this tree for its element at position defined by \coqinline{il'}. The case $i_{d - 1} = 1$ and $n_{d - 1} = 0$ is impossible on valid input.\\% bad

@@

All that is left now is to check whether \coqinline{i} is in bounds (to get a functions that works on and \coqinline{i}) and to convert it into digits.

@@

\begin{center}
    \begin{tabular}{|c|c||c|c|}
        \hline
        \textbf{First operand} & \textbf{Second operand} & \textbf{Carry} & \textbf{Result} \\
        \hline
        $0$ & $0$ & $0$ & $0$ \\
        $0$ & $1$ & $0$ & $1$ \\
        $1$ & $0$ & $0$ & $1$ \\
        $1$ & $1$ & $1$ & $0$ \\
        \hline
    \end{tabular}
\end{center}

@@

We do not try to maintain the minimum possible depth, so the head of the digital list can become empty.\\% bad

@@

To fix that, we could just define predicates that asserts that the data structure are valid, and that option was also tried. But here we will explore another way to achieve what we want: dependent types, which allows to store proofs of data structure validity inside data structures themselves.

@@

@@

The definitions of a digital list and a concrete digital list are similar to the ones seen before, but now, as promised, contain requirements that make data structures of these types always valid.

@@

\begin{coq}
Check concrete_digital_list_to_list : forall {A} {r}, concrete_digital_list A r -> list A.
Check concrete_digital_list_empty : forall {A} {r}, concrete_digital_list A r.
Check concrete_digital_list_length : forall {A} {r}, concrete_digital_list A r -> nat.
Check concrete_digital_list_nth : forall {A} {r} (i : nat), concrete_digital_list A r -> option A.
Check concrete_digital_list_update : forall {A} {r}, nat -> A -> concrete_digital_list A r ->
  option (concrete_digital_list A r).
Check concrete_digital_list_push : forall {A} {r}, A -> concrete_digital_list A r -> concrete_digital_list A r.
Check concrete_digital_list_pop : forall {A} {r}, concrete_digital_list A r ->
  option (concrete_digital_list A r * A).
\end{coq}

@@

\vspace{0.5cm}% bad

@@

Inductive leaf_tree_complete {A} r : nat -> leaf_tree A -> Prop :=
  | LeafTreeCompleteLeaf : forall x, leaf_tree_complete r 0 (LeafTreeLeaf x)
  | LeafTreeCompleteInternalNode :
      forall {d} a,
      array_length a = r ->
      List.Forall (leaf_tree_complete r d) (array_to_list a) ->
      leaf_tree_complete r (S d) (LeafTreeInternalNode a).

Inductive digital_list {A} :=
  | DigitalListNil : digital_list
  | DigitalListCons : array (leaf_tree A) -> digital_list -> digital_list.

@@

\begin{coq}
Extract Inductive array => ``array'' [ ``(assert false)'' ] ``(assert false)''.
Extract Constant array_to_list => ``fun _ a -> Array.to_list a''.
Extract Inlined Constant array_empty => ``[||]''.
Extract Constant array_single => ``fun x -> [|x|]''.
Extract Constant array_nth => ``
  fun _ i a ->
    try Some a.(i)
    with Invalid_argument _ -> None
''.
Extract Constant array_update => ``
  fun _ i x a ->
    let a0 = Array.copy a in
    try
      a0.(i) <- x;
      Some a0
    with Invalid_argument _ -> None
''.
Extract Constant array_push => ``
  fun n x a ->
    let a0 = Array.make (n + 1) x in
    Array.blit a 0 a0 0 n;
    a0
''.
Extract Constant array_pop => ``fun n a -> (Array.sub a 0 n, a.(n))''.
\end{coq}

@@

Even thought this way of measuring data structure size is very crude, we can still still the the extraction of the dependent version works faster and uses less memory than the non-dependent one. The reason is that the non-dependent version needs more indirections when storing leaf trees, which are bad for performance and memory footprint.

@@

@@

\begin{center}
    \begin{tabular}{|l|r|}
        \hline
        Common code & 426 lines \\
        \hline
        \coqinlinelight{r}-ary dependent version & 1030 lines \\
        \hline
        \coqinlinelight{r}-ary non-dependent version & 1107 lines \\
        \hline
    \end{tabular}
\end{center}

@@

>>>>

\hypersetup{colorlinks=true, urlcolor=blue}

@@

\usepackage{array}

@@

\usepackage{pgfplots}
\pgfplotsset{compat = 1.17}

@@

We know that for each natural number $n$ there are $d \in \mathbb{N}$ and $n_{d - 1}, n_{d - 2}, ..., n_1, n_0 \in \{0, 1\}$ such that $n = n_{d - 1} 2^{d - 1} + n_{d - 2} 2^{d - 2} + ... + n_1 2^1 + n_0 2^0$. The function \coqinline{to_digits : forall (d n : nat), list nat} gives the list of length \coqinline{d} consisting of these binary digits of \coqinline{n} in big-endian ordering: for example, \coqinline{to_digits 8 73 ~> [0; 1; 0; 0; 1; 0; 0; 1]} (where ``\nospaceevaluatesto'' means ``evaluates to''). The inverse operation is \coqinline{of_digits : forall (d : nat) (nl : list nat), nat}: for example, \coqinline{of_digits 8 [0; 1; 0; 0; 1; 0; 0; 1] ~> 73}.

@@

Here is an example of a digit list for $n = 5$, so we can chose $d = 3$ (the minimum number of bits needed to represent $5$), $n_2$ = 1, $n_1 = 0$, $n_0 = 1$, which stores values 0, 1, 2, 3, 4:

@@

So in the case $i_{d - 1} = n_{d - 1} = 0$ (that is, \coqinline{il} starts with 0 and there is no tree in this position) and in the case $i_{d - 1} = n_{d - 1} = 1$ (that is, \coqinline{il} starts with 1 and there is a tree in this position), we recurse to our tail. Otherwise, in the case $i_{d - 1} = 0$ and $n_{d - 1} = 1$ (that is, \coqinline{il} starts with 0 and there is a tree in this position), we ask this tree for its element at position defined by \coqinline{il'}. The case $i_{d - 1} = 1$ and $n_{d - 1} = 0$ is impossible on valid input. \pagebreak% bad

@@

All that is left now is to check whether \coqinline{i} is in bounds (to get a function that works on all \coqinline{i}) and to convert it into digits.

@@

\begin{center}
    \newcolumntype{M}{>{\centering\arraybackslash}m{45pt}}

    \begin{tabular}{|M|M||M|M|}
        \hline
        \textbf{First operand} & \textbf{Second operand} & \textbf{Carry} & \textbf{Result} \\
        \hline
        $0$ & $0$ & $0$ & $0$ \\
        $0$ & $1$ & $0$ & $1$ \\
        $1$ & $0$ & $0$ & $1$ \\
        $1$ & $1$ & $1$ & $0$ \\
        \hline
    \end{tabular}
\end{center}

@@

We do not try to maintain the minimum possible depth, so the head of the digital list can become empty. \pagebreak% bad

@@

To fix that, we could just define predicates that asserts that the data structure are valid, and that option was also tried. But here we will explore another way to achieve what we want: dependent types, which allows to store proofs of data structure invariants inside data structures themselves.

@@

We will use \coqinline{r}-ary complete trees instead of binary complete trees. Also we will store between $0$ and \coqinline{r - 1} of these trees in each position of a digital tree, and not $0$ or $1$ as before (represented by the \coqinline{option} type).

@@

The definitions of a digital list and a concrete digital list are similar to the ones seen before, but now, as promised, contain requirements that make data structures of these types always valid. Here \coqinline{k} is the number of trees at a position of digital trees, and it, as said before, will be a natural number strictly less than \coqinline{r}.

@@

\begin{coq}
Check concrete_digital_list_to_list : forall {A} {r}, concrete_digital_list A r -> list A.
Check concrete_digital_list_empty : forall {A} {r}, concrete_digital_list A r.
Check concrete_digital_list_length : forall {A} {r}, concrete_digital_list A r -> nat.
Check concrete_digital_list_nth : forall {A} {r} (i : nat), concrete_digital_list A r -> option A.
Check concrete_digital_list_update : forall {A} {r}, nat -> A -> concrete_digital_list A r ->
  option (concrete_digital_list A r).
Check concrete_digital_list_push : forall {A} {r}, A -> concrete_digital_list A r -> concrete_digital_list A r.
Check concrete_digital_list_pop : forall {A} {r}, concrete_digital_list A r ->
  option (concrete_digital_list A r * A).
\end{coq} \pagebreak% bad

@@

@@

Inductive leaf_tree_complete {A} r : nat -> leaf_tree A -> Prop :=
  | LeafTreeCompleteLeaf : forall x, leaf_tree_complete r 0 (LeafTreeLeaf x)
  | LeafTreeCompleteInternalNode :
      forall {d} a,
      array_length a = r ->
      List.Forall (leaf_tree_complete r d) (array_to_list a) ->
      leaf_tree_complete r (S d) (LeafTreeInternalNode a).
\end{coq}
\pagebreak% bad
\begin{coq}
Inductive digital_list {A} :=
  | DigitalListNil : digital_list
  | DigitalListCons : array (leaf_tree A) -> digital_list -> digital_list.

@@

\begin{coq}
Extract Inductive array => ``array'' [ ``(assert false)'' ] ``(assert false)''.
Extract Constant array_to_list => ``fun _ a -> Array.to_list a''.
Extract Inlined Constant array_empty => ``[||]''.
Extract Constant array_single => ``fun x -> [|x|]''.
Extract Constant array_nth => ``
  fun _ i a ->
    try Some a.(i)
    with Invalid_argument _ -> None
''.
Extract Constant array_update => ``
  fun _ i x a ->
    let a0 = Array.copy a in
    try
      a0.(i) <- x;
      Some a0
    with Invalid_argument _ -> None
''.
\end{coq}
\pagebreak% bad
\begin{coq}
Extract Constant array_push => ``
  fun n x a ->
    let a0 = Array.make (n + 1) x in
    Array.blit a 0 a0 0 n;
    a0
''.
Extract Constant array_pop => ``fun n a -> (Array.sub a 0 n, a.(n))''.
\end{coq}

@@

Even thought this way of measuring data structure size is very crude, we can still the the extraction of the dependent version works faster and uses less memory than the non-dependent one. The reason is that the non-dependent version needs more indirections when storing leaf trees, which are bad for performance and memory footprint.

@@

For the choice of \coqinline{r}, here is how much time executing the code above (but with $1000000$ as digital lists length) for the dependent version takes for different values of \coqinline{r} (logarithmic scale):

\begin{center}
    \begin{tikzpicture}
        \begin{axis}[xlabel = \coqinline{r}, ylabel = {Time (ms, lower is better)}, xlabel shift = -30, ylabel shift = -40, xmode = log, log basis x = 2, width = \textwidth, height = 10cm]
            \addplot[mark = none] table [x = r, y = time, col sep = comma] {../extraction/rs_output.csv};
        \end{axis}
    \end{tikzpicture}
\end{center}

We can see that the optimal radix in this particular size is around $32$. We can also remark that the binary case is many times slower, and so generalizing the data structure to different radixes can actually be useful for some scenarios.

@@

\begin{center}
    \begin{tabular}{|l|r|}
        \hline
        \href{https://github.com/afdw/digital_list/blob/main/theories/Utils.v}{Common code} & 426 lines of code \\
        \hline
        \href{https://github.com/afdw/digital_list/tree/main/theories/Dep}{\coqinlinelight{r}-ary dependent version} & 1030 lines of code \\
        \hline
        \href{https://github.com/afdw/digital_list/tree/main/theories/NonDep}{\coqinlinelight{r}-ary non-dependent version} & 1107 lines of code \\
        \hline
    \end{tabular}
\end{center}

@@

The source code of the development is available at \url{https://github.com/afdw/digital_list}.
